---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

Load the dataset specifying the class for the columns

```{r, echo = TRUE}
activityData <- read.csv("activity.csv", header = TRUE, colClasses = c("integer", "Date", "integer"))
```

## What is the mean total number of steps taken per day?
### Calculate the total number of steps taken per day
```{r, echo = TRUE}
stepsPerDay <- aggregate(steps ~ date, activityData, sum)
```
This is a small subset of the result

```{r, echo = TRUE}
head(stepsPerDay)
```

### Make a histogram of the total number of steps taken each day
```{r, echo = TRUE}
hist(stepsPerDay$steps, main = "Histogram of Steps per Day", xlab = "Steps per Day", col = "lightblue", breaks = 10)
```

### Calculate and report the mean and median of the total number of steps taken per day
```{r, echo = TRUE}
meanSteps <- mean(stepsPerDay$steps, na.rm = TRUE)
medianSteps <- median(stepsPerDay$steps, na.rm = TRUE)
```

Mean of the total number of steps taken pery day: **`r format(meanSteps, scientific = FALSE)`**

Median of the total number of steps taken pery day: **`r format(medianSteps, scientific = FALSE)`**

## What is the average daily activity pattern?
### Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r, echo = TRUE}
avgStepsPerInterval <- aggregate(steps ~ interval, activityData, mean, na.rm = TRUE)

plot(x = avgStepsPerInterval$interval, y = avgStepsPerInterval$steps, 
     type = "l", main = "Average number of steps taken per 5-minute interval", 
     xlab = "5-minute interval", ylab = "Average number of steps", col = "blue", lwd = 2)
```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r, echo = TRUE}
maxAvgSteps <- which.max(avgStepsPerInterval$steps)
maxInterval <- avgStepsPerInterval[maxAvgSteps,]$interval
```
5-minute interval containing the maximum number of steps: **`r maxInterval`**

## Imputing missing values
### Calculate and report the total number of missing values in the dataset
```{r, echo = TRUE}
totalNA <- sum(is.na(activityData))
```

Total number of missing values: **`r totalNA`**

### Devise a strategy for filling in all of the missing values in the dataset
Use the mean for the corresponding 5-minute interval rounded to the nearest integer

To do that the previous **avgStepsPerInterval** data frame is used, which contains the average number of steps taken per interval, averaged across all days

**newData** is a dataset equal to the original dataset but with the missing values filled in
```{r, echo = TRUE}
newData <- activityData
naIndex <- which(is.na(newData))
for (i in naIndex) {
  newData[i,]$steps <- round(avgStepsPerInterval[(avgStepsPerInterval$interval == newData[i,]$interval),]$steps)
}
```

### Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day
```{r, echo = TRUE}
stepsPerDay <- aggregate(steps ~ date, newData, sum)
hist(stepsPerDay$steps, main = "Histogram of Steps per Day", xlab = "Steps per Day", col = "blue", breaks = 10)

meanSteps2 <- mean(stepsPerDay$steps, na.rm = TRUE)
medianSteps2 <- median(stepsPerDay$steps, na.rm = TRUE)
```

### Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
The mean and median differ from the previous ones. The impact of imputing missing values is that **both the mean and the median slightly decrease**

Original mean: **`r format(meanSteps, scientific = FALSE)`**

Mean after imputing missing data: **`r format(meanSteps2, scientific = FALSE)`**

Original median: **`r format(medianSteps, scientific = FALSE)`**

Median after imputing missing data: **`r format(medianSteps2, scientific = FALSE)`**

## Are there differences in activity patterns between weekdays and weekends?
### Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day
```{r, echo = TRUE}
newData$day <- as.factor(ifelse(weekdays(newData$date) %in% c("Saturday", "Sunday"), "weekend", "weekday"))
```

### Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)
```{r, echo = TRUE}
library(lattice)
avgSteps <- aggregate(steps ~ interval + day, newData, mean)
xyplot(steps ~ interval | day, data = avgSteps, type = "l", layout = c(1, 2))
```