---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

Load the dataset specifying the class for the columns

```{r, echo = TRUE}
activityData <- read.csv("activity.csv", header = TRUE, colClasses = c("integer", "Date", "integer"))
```

## What is mean total number of steps taken per day?
### Calculate the total number of steps taken per day
```{r, echo = TRUE}
stepsPerDay <- aggregate(steps ~ date, activityData, sum)
```

### Make a histogram of the total number of steps taken each day
```{r, echo = TRUE}
hist(stepsPerDay$steps, main = "Histogram of Steps per Day", xlab = "Steps per Day", col = "lightblue", breaks = 10)
```

### Calculate and report the mean and median of the total number of steps taken per day
```{r, echo = TRUE}
meanSteps <- mean(stepsPerDay$steps, na.rm = TRUE)
medianSteps <- median(stepsPerDay$steps, na.rm = TRUE)
```

Mean of the total number of steps taken pery day: **`r format(meanSteps, scientific = FALSE)`**

Median of the total number of steps taken pery day: **`r format(medianSteps, scientific = FALSE)`**

## What is the average daily activity pattern?
### Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r, echo = TRUE}
avgStepsPerInterval <- aggregate(steps ~ interval, activityData, mean, na.rm = TRUE)

plot(x = avgStepsPerInterval$interval, y = avgStepsPerInterval$steps, 
     type = "l", main = "Average number of steps taken per 5-minute interval", 
     xlab = "5-minute interval", ylab = "Average number of steps", col = "blue", lwd = 2)
```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r, echo = TRUE}
maxAvgSteps <- which.max(avgStepsPerInterval$steps)
maxInterval <- avgStepsPerInterval[maxAvgSteps,]$interval
```
5-minute interval containing the maximum number of steps: **`r maxInterval`**

## Imputing missing values
### Calculate and report the total number of missing values in the dataset
```{r, echo = TRUE}
totalNA <- sum(is.na(activityData))
```

Total number of missing values: **`r totalNA`**

### Devise a strategy for filling in all of the missing values in the dataset
Use the mean for that 5-minute interval rounded to the nearest integer
newData is a dataset equal to the original dataset but with the missing values filled in
```{r, echo = TRUE}
newData <- activityData
naIndex <- which(is.na(newData))
for (i in naIndex) {
  newData[i,]$steps <- round(avgStepsPerInterval[(avgStepsPerInterval$interval == newData[i,]$interval),]$steps)
}
```

### Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day
```{r, echo = TRUE}
stepsPerDay <- aggregate(steps ~ date, newData, sum)
hist(stepsPerDay$steps, main = "Histogram of Steps per Day", xlab = "Steps per Day", col = "blue", breaks = 10)

meanSteps2 <- mean(stepsPerDay$steps, na.rm = TRUE)
medianSteps2 <- median(stepsPerDay$steps, na.rm = TRUE)
```

### Do these values differ from the estimates from the first part of the assignment?
Original mean: **`r format(meanSteps, scientific = FALSE)`**

Mean after imputing missing data: **`r format(meanSteps2, scientific = FALSE)`**

Original median: **`r format(medianSteps, scientific = FALSE)`**

Median after imputing missing data: **`r format(medianSteps2, scientific = FALSE)`**

### What is the impact of imputing missing data on the estimates of the total daily number of steps?
Both the mean and the median slightly decrease

## Are there differences in activity patterns between weekdays and weekends?
### Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day
```{r, echo = TRUE}
newData$day <- as.factor(ifelse(weekdays(newData$date) %in% c("Saturday", "Sunday"), "weekend", "weekday"))
```

### Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)
```{r, echo = TRUE}
library(lattice)
avgSteps <- aggregate(steps ~ interval + day, newData, mean)
xyplot(steps ~ interval | day, data = avgSteps, type = "l", layout = c(1, 2))
```